### 2023-09-05 proofs/tex/equiv-alg-lemmas.tex
Rearrange proofs

\begin{lemma} [Declarative Equivalence is invariant under bijections]
  \label{lemma:decl-equiv-bijection}
  Suppose $[[μ]]$ is a bijection $[[μ : varset1 ↔ varset2]]$, then
  \begin{itemize}
    \item[$+$] $[[iP1 ≈ iP2]]$ implies $[[ [μ]iP1 ≈ [μ]iP2 ]]$,
      and there exists an inference tree of $[[ [μ]iP1 ≈ [μ]iP2 ]]$
      with the same shape as the one inferring $[[iP1 ≈ iP2]]$;
    \item[$-$] $[[iN1 ≈ iN2]]$ implies $[[ [μ]iN1 ≈ [μ]iN2 ]]$, 
      and there exists an inference tree of $[[ [μ]iN1 ≈ [μ]iN2 ]]$
      with the same shape as the one inferring $[[iN1 ≈ iN2]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  We prove it by induction on $[[iP1 ≈ iP2]]$ and mutually, on $[[iN1 ≈ iN2]]$.
  Let us consider the last rule used in the derivation.
  \begin{caseof}
    \item \ruleref{\ottdruleEOneForallLabel}\\
      Then we decompose $[[iN1]]$ as $[[∀pas1.iM1]]$ and $[[iN2]]$ as $[[∀pas2.iM2]]$,
      where $[[iM1]]$ and $[[iM2]]$ do not start with $[[∀]]$-quantifiers.
      where $[[|pas1| + |pas2| > 0]]$.
      By convention, let us assume that
      $[[pas1]]$ and $[[pas2]]$ are disjoint form 
      $[[varset2]]$ and $[[varset1]]$.

      By inversion, $[[{pas1} ∩ fv iM2 = ∅]]$ and 
      $[[iM1 ≈ [μ'] iM2]]$ for some bijection
      $[[μ' : ({pas2} ∩ fv iM2) ↔ ({pas1} ∩ fv iM1)]]$.
      Then let us apply the induction hypothesis to
       $[[iM1 ≈ [μ'] iM2]]$ to obtain $[[ [μ]iM1 ≈ [μ] [μ'] iM2 ]]$ 
      inferred by the tree of the same shape as 
      $[[ iM1 ≈ [μ'] iM2 ]]$.

      Notice that $[[ [μ]iM1 ]]$ and $[[ [μ]iM2 ]]$ 
      do not start with $[[∀]]$, 
      That is $[[ [μ]∀pas1.iM1 ≈ [μ]∀pas2.iM2]]$, 
      rewritten as $[[ ∀pas1.[μ]iM1 ≈ ∀pas2.[μ]iM2]]$,
      can be inferred by \ruleref{\ottdruleEOneForallLabel}:
      \begin{enumerate}
        \item $[[{pas1}]]$ is disjoint from 
        $[[varset2 ∪ fv iM2 ⊆ fv [μ]iM2 ]]$;
        \item $[[ [μ]iM1 ≈ [μ'] [μ] iM2 ]]$
          because $[[ [μ'] [μ] iM2 = [μ] [μ'] iM2 ]]$
          (by \cref{corollary:subst-composition-commutativity}:
          $[[μ' : ({pas2} ∩ fv iM2) ↔ ({pas1} ∩ fv iM1)]]$, 
          $[[μ : varset1 ↔ varset2]]$,
          $[[varset1]]$ is disjoint from 
          $[[pas2]]$ and $[[pas1]]$; 
          $[[pas2]]$ is disjoint from $[[varset1]]$ and $[[varset2]]$)
      \end{enumerate}
      Notice that it is the same rule as the one inferring 
      $[[iN1 ≈ iN2]]$, and thus, the shapes of the trees are the same.

    \item \ruleref{\ottdruleEOneNVarLabel}\\
      Then $[[iN1]] = [[iN2]] = [[α⁻]]$, and the required
      $[[ [μ]α⁻ = [μ]α⁻ ]]$ is also inferred by
      \ruleref{\ottdruleEOneNVarLabel}, since 
      $[[ [μ]α⁻ ]]$ is a variable.
    \item \ruleref{\ottdruleEOneArrowLabel}\\
      Then we are proving 
      that $[[iP1 → iM1 ≈ iP2 → iM2]]$ implies
      $[[ [μ](iP1 → iM1) ≈ [μ](iP2 → iM2) ]]$ 
      (preserving the tree structure).

      By inversion, we have $[[iP1 ≈ iP2]]$ and $[[iM1 ≈ iM2]]$,
      and thus, by the induction hypothesis,
      $[[ [μ]iP1 ≈ [μ]iP2 ]]$ and $[[ [μ]iM1 ≈ [μ]iM2 ]]$.
      Then $[[ [μ](iP1 → iM1) ≈ [μ](iP2 → iM2) ]]$, 
      or in other words, $[[ [μ]iP1 → [μ]iM1 ≈ [μ]iP2 → [μ]iM2 ]]$,
      is inferred by the same rule---\ruleref{\ottdruleEOneArrowLabel}.
    \item \ruleref{\ottdruleEOneShiftULabel}
      This case is done by similar congruent arguments as
      the previous one.
    \item The positive cases are proved symmetrically.
    \end{caseof}
\end{proof}

\begin{lemma}[Declarative equivalence is transitive]
  \hfill
  \label{lemma:decl-equiv-transitivity}
  \begin{itemize}
    \item[$+$] if $[[iP1 ≈ iP2]]$ and $[[iP2 ≈ iP3]]$ then $[[iP1 ≈ iP3]]$,
    \item[$-$] if $[[iN1 ≈ iN2]]$ and $[[iN2 ≈ iN3]]$ then $[[iN1 ≈ iN3]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  We prove it by
  $\size{[[iP1 ≈ iP2]]} + \size{[[iP2 ≈ iP3]]}$
  and mutually, $\size{[[iN1 ≈ iN2]]} + \size{[[iN2 ≈ iN3]]}$, where 
  by size, we mean the size of the nodes in the corresponding inference tree.

  \begin{caseof}
    \item First, let us consider the case when either 
      $[[iN1 ≈ iN2]]$ or $[[iN2 ≈ iN3]]$ is inferred by 
      \ruleref{\ottdruleEOneForallLabel}.
      Let us decompose $[[iN1]]$, $[[iN2]]$, and $[[iN3]]$ as follows:
      $[[iN1 = ∀pas1.iM1]]$,
      $[[iN2 = ∀pas2.iM2]]$, and
      $[[iN3 = ∀pas3.iM3]]$.

      Then by inversion of $[[∀pas1.iM1 ≈ ∀pas2.iM2]]$
      (or if $[[pas1]]$ and $[[pas2]]$ are both empty, 
      by assumption) : 
      \begin{enumerate}
        \item $[[{pas1} ∩ fv iM2 = ∅]]$ and
        \item there exists a bijection on variables 
          $[[mu1 : ({pas2} ∩ fv iM2) ↔ ({pas1} ∩ fv iM1)]]$
          such that $[[iM1 ≈ [mu1] iM2]]$.
      \end{enumerate}

      Analogously, $[[∀pas1.iM1 ≈ ∀pas2.iM2]]$
      implies:
      \begin{enumerate}
        \item $[[{pas2} ∩ fv iM3 = ∅]]$ and
        \item $[[iM2 ≈ [mu2] iM3]]$ for some bijection 
          $[[mu2 : ({pas3} ∩ fv iM3) ↔ ({pas2} ∩ fv iM2)]]$.
      \end{enumerate}

      Notice that either $[[iM1 ≈ [mu1] iM2]]$
      is inferred by a proper sub-tree of $[[∀pas1.iM1 ≈ ∀pas2.iM2]]$
      or $[[iM2 ≈ [mu2] iM3]]$ is inferred by a proper sub-tree of 
      $[[∀pas2.iM2 ≈ ∀pas3.iM3]]$.

      Then by \cref{lemma:decl-equiv-bijection}, 
      $[[ [mu1]iM2 ≈ [mu1 ○ mu2] iM3]]$ and moreover, 
      $\size{[[ [mu1]iM2 ≈ [mu1 ○ mu2] iM3]]} = \size{[[iM2 ≈ [mu2] iM3]]}$.

      Since at least one of the trees inferring 
      $[[iM1 ≈ [mu1] iM2]]$ and $[[iM2 ≈ [mu2] iM3]]$
      is a proper sub-tree of the corresponding original tree,
      $\size{[[iM1 ≈ [mu1] iM2]]} + \size{[[iM2 ≈ [mu2] iM3]]} <
      \size{[[∀pas1.iM1 ≈ ∀pas2.iM2]]} + \size{[[∀pas2.iM2 ≈ ∀pas3.iM3]]}$,
      i.e., the induction hypothesis is applicable.

      By the induction hypothesis, $[[iM1 ≈ [mu1 ○ mu2] iM3]]$.
      Where $[[mu1 ○ mu2]]$ is a bijection on variables
      $[[mu1 ○ mu2 : ({pas3} ∩ fv iM3) ↔ ({pas1} ∩ fv iM1)]]$.
      Then $[[∀pas1.iM1 ≈ ∀pas3.iM3]]$ by \ruleref{\ottdruleEOneForallLabel}.

      Once this case has been considered, 
      we can assume that neither $[[iN1 ≈ iN2]]$ nor $[[iN2 ≈ iN3]]$
      is inferred by \ruleref{\ottdruleEOneForallLabel}.

    \item $[[iN1 ≈ iN2]]$ is inferred by 
      \ruleref{\ottdruleEOneNVarLabel}\\
      Then $[[iN1]] = [[iN2]] = [[α⁻]]$, and thus,  
      $[[iN1 ≈ iN3]]$ holds since $[[iN2 ≈ iN3]]$.

    \item $[[iN1 ≈ iN2]]$ is inferred by 
      \ruleref{\ottdruleEOneArrowLabel}\\
      Then $[[iN1]] = [[iP1 → iM1]]$ and $[[iN2]] = [[iP2 → iM2]]$,
      and by inversion, $[[iP1 ≈ iP2]]$ and $[[iM1 ≈ iM2]]$.

      Moreover, since $[[iN3]]$ does not start with $[[∀]]$,
      $[[iN2 ≈ iN3]]$ is also inferred by 
      \ruleref{\ottdruleEOneArrowLabel},
      which means that $[[iN3]] = [[iP3 → iM3]]$,
      $[[iP2 ≈ iP3]]$, and $[[iM2 ≈ iM3]]$.

      Then by the induction hypothesis, 
      $[[iP1 ≈ iP3]]$ and $[[iM1 ≈ iM3]]$, and thus, 
      $[[iP1 → iM1 ≈ iP3 → iM3]]$ by \ruleref{\ottdruleEOneArrowLabel}.
    \item $[[iN1 ≈ iN2]]$ is inferred by
      \ruleref{\ottdruleEOneArrowLabel}\\
      For this case, the reasoning is the same as for the previous one.
    \item The positive cases are proved symmetrically.
  \end{caseof}
\end{proof}

@@

\begin{lemma}[Type well-formedness is invariant under equivalence]
  \label{lemma:wf-equiv}
  Mutual subtyping implies declarative equivalence.
  \begin{itemize}
  \item[$+$] if $[[iP ≈ iQ]]$ then $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$,
  \item[$-$] if $[[iN ≈ iM]]$ then $[[Γ ⊢ iN]] \iff [[Γ ⊢ iM]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  We prove it by induction on $[[iP ≈ iQ]]$ and mutually, on $[[iN ≈ iM]]$.
  Let us consider the last rule used in the derivation.
  \begin{caseof}
    \item \ruleref{\ottdruleEOneNVarLabel}, 
      that is $[[iN ≈ iM]]$ has shape $[[α⁻ ≈ α⁻]]$.\\
      Than $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$ is rewritten as $[[Γ ⊢ α⁻]] \iff [[Γ ⊢ α⁻]]$,
      which holds trivially.
    \item \ruleref{\ottdruleEOneShiftULabel},
      that is $[[iN ≈ iM]]$ has shape $[[↑iP ≈ ↑iQ]]$.\\
      By inversion, $[[iP ≈ iQ]]$, and by the induction hypothesis,
      $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$. 
      Also notice that $[[Γ ⊢ ↑iP]] \iff [[Γ ⊢ iP]]$ 
      and $[[Γ ⊢ ↑iQ]] \iff [[Γ ⊢ iQ]]$ by inversion and \ruleref{\ottdruleWFTShiftULabel}.
      This way, $[[Γ ⊢ ↑iP]] \iff [[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]] \iff [[Γ ⊢ ↑iQ]]$.
    \item \ruleref{\ottdruleEOneArrowLabel},
      that is $[[iN ≈ iM]]$ has shape $[[iP → iN' ≈ iQ → iM']]$.\\
      Then by inversion, $[[iP ≈ iQ]]$ and $[[iN' ≈ iM']]$, 
      and by the induction hypothesis, $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$ and $[[Γ ⊢ iN']] \iff [[Γ ⊢ iM']]$.
      $
      \begin{aligned}[t]
        [[Γ ⊢ iP → iN']]  &\iff [[Γ ⊢ iP]] \text{ and } [[Γ ⊢ iN']] 
                          && \text{by inversion and \ruleref{\ottdruleWFTArrowLabel}}\\
                          &\iff [[Γ ⊢ iQ]] \text{ and } [[Γ ⊢ iM']]
                          && \text{as noted above}\\
                          &\iff [[Γ ⊢ iQ → iM']]
                          && \text{by \ruleref{\ottdruleWFTArrowLabel} and inversion}\\
      \end{aligned}
      $\\
    \item \ruleref{\ottdruleEOneForallLabel},
      that is $[[iN ≈ iM]]$ has shape $[[∀pas.iN' ≈ ∀pbs.iM']]$.\\
      By inversion, $[[∀pas.iN' ≈ ∀pbs.iM']]$ means that
      $[[{pas} ∩ fv iM = ∅]]$ and 
      that there exists a bijection on variables 
      $[[mu : ({pbs} ∩ fv iM') ↔ ({pas} ∩ fv iN')]]$
      such that $[[iN' ≈ [mu] iM']]$.

      By inversion and \ruleref{\ottdruleWFTForallLabel},
      $[[Γ ⊢ ∀pas.iN']]$ is equivalent to $[[Γ, pas ⊢ iN']]$,
      and by \cref{lemma:wf-ctxt-equiv}, it is equivalent to 
      $[[Γ,  ({pas} ∩ fv iN') ⊢ iN']]$,
      which, by the induction hypothesis, is equivalent to
      $[[Γ,  ({pas} ∩ fv iN') ⊢ [mu]iM']]$.

      Analogously, $[[Γ ⊢ ∀pbs.iM']]$ is equivalent to
      $[[Γ, ( {pbs} ∩ fv iM') ⊢ iM']]$.
      By \cref{lemma:wf-subst}, it implies 
      $[[Γ,  ({pas} ∩ fv iM') ⊢ [mu]iM']]$. And vice versa, 
      $[[Γ,  ({pas} ∩ fv iM') ⊢ [mu]iM']]$ implies  
      $[[Γ,  ({pbs} ∩ fv iM') ⊢ [mu-1][mu]iM']]$.

      This way, both $[[Γ ⊢ ∀pas.iN']]$ and $[[Γ ⊢ ∀pbs.iM']]$ are equivalent to
      $[[Γ,  ({pas} ∩ fv iN') ⊢ [mu]iM']]$. 
    \item For the cases of the positive types, the proofs are symmetric.
  \end{caseof}
\end{proof}

\begin{lemma}[Algorithmic type well-formedness is invariant under equivalence]
  \label{lemma:wf-equiv-algo}
  Mutual subtyping implies declarative equivalence.
  \begin{itemize}
  \item[$+$] if $[[uP ≈ uQ]]$ then $[[Γ; Ξ ⊢ uP]] \iff [[Γ; Ξ ⊢ uQ]]$,
  \item[$-$] if $[[uN ≈ uM]]$ then $[[Γ; Ξ ⊢ uN]] \iff [[Γ; Ξ ⊢ uM]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  The proof coincides with the proof of \cref{lemma:wf-equiv},
  and adds two cases for equating two positive or two negative metavariables,
  which must be equal by inversion, and thus, 
  $[[Γ; Ξ ⊢ α̂±]] \iff [[Γ; Ξ  ⊢ α̂±]]$ holds trivially.
\end{proof}

\begin{corollary}[Normalization preserves well-formedness]
  \label{corollary:wf-nf}
  \hfill
  \begin{itemize}
  \item[$+$] $[[Γ ⊢ iP]] \iff [[Γ ⊢ nf(iP)]]$,
  \item[$-$] $[[Γ ⊢ iN]] \iff [[Γ ⊢ nf(iN)]]$
  \end{itemize}
\end{corollary}
\begin{proof}
  Immediately from \cref{lemma:wf-equiv,lemma:normalization-soundness}.
\end{proof}

\begin{lemma}[Normalization preserves substitution signature]
  \label{lemma:norm-subst-sig}
  Suppose that $[[σ]]$ is a substitution, $[[Γ1]]$ and $[[Γ2]]$ are contexts.
  Then $[[Γ2 ⊢ σ : Γ1]]$ implies $[[Γ2 ⊢ nf(σ) : Γ1]]$.
\end{lemma}
\begin{proof}
  Suppose that $[[α± ∊ Γ1]]$. 
  Then by \cref{corollary:wf-nf}, $[[Γ2 ⊢ nf([σ]α±)]] = [[ [nf(σ)]α± ]]$ 
  is equivalent to $[[Γ2 ⊢ [σ]α±]]$.

  Suppose that $[[α± ∉ Γ1]]$. 
  $[[Γ2 ⊢ σ : Γ1]]$ means that $[[ [σ]α± = α± ]]$, 
  and then $[[ [nf(σ)]α± ]] = [[nf([σ]α±)]] = [[nf(α±)]] = [[α±]]$.
\end{proof}

@@

\begin{corollary}[Normalization preserves well-formedness of substitution]
  \label{corollary:wf-s-nf}
  \hfill \\
   $[[Γ2 ⊢ σ : Γ1]] \iff [[Γ2 ⊢ nf(σ) : Γ1]]$
\end{corollary}
\begin{proof}
  Let us prove the forward direction.
  Suppose that $[[α± ∊ Γ1]]$.  Let us show that $[[Γ2 ⊢ [nf(σ)]α±]]$.
  By the definition of substitution normalization,
  $[[ [nf(σ)]α± = nf([σ]α±) ]]$. Then by \cref{corollary:wf-nf-algo},
  to show $[[Γ2 ⊢ nf([σ]α±)]]$, it suffices to show $[[Γ2 ⊢ [σ]α±]]$,
  which holds by the assumption $[[Γ2 ⊢ σ : Γ1]]$.

  The backward direction is proved analogously.
\end{proof}

@@

\begin{lemma}[Soundness of equivalence]
  \label{lemma:equiv-soundness}
  Declarative equivalence implies mutual subtyping.
  \begin{itemize}
    \item[$+$] if $[[Γ ⊢ iP]]$, $[[Γ ⊢ iQ]]$, and $[[iP ≈ iQ]]$ then $[[Γ ⊢ iP ≈ iQ]]$,
    \item[$-$] if $[[Γ ⊢ iN]]$, $[[Γ ⊢ iM]]$, and $[[iN ≈ iM]]$ then $[[Γ ⊢ iN ≈ iM]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  We prove it by mutual induction on $[[iP ≈ iQ]]$ and $[[iN ≈ iM]]$.
  \begin{caseof}
    \item $[[a⁻ ≈ a⁻]]$\\
      Then $[[Γ ⊢ a⁻ ≤ a⁻]]$ by \ruleref{\ottdruleDOneNVarLabel},
      which immediately implies $[[Γ ⊢ a⁻ ≈ a⁻]]$ by \ruleref{\ottdruleDOneNDefLabel}.

    \item $[[↑iP ≈ ↑iQ]]$\\
      Then by inversion of \ruleref{\ottdruleDOneShiftULabel},
      $[[iP ≈ iQ]]$, and from the induction hypothesis, $[[Γ ⊢ iP ≈ iQ]]$,
      and (by symmetry) $[[Γ ⊢ iQ ≈ iP]]$.

      When \ruleref{\ottdruleDOneShiftULabel} is applied to $[[Γ ⊢ iP ≈ iQ]]$,
      it gives us $[[Γ ⊢ ↑iP ≤ ↑iQ]]$; when it is applied to $[[Γ ⊢ iQ ≈ iP]]$,
      we obtain $[[Γ ⊢ ↑iQ ≤ ↑iP]]$. Together, it implies $[[Γ ⊢ ↑iP ≈ ↑iQ]]$.

    \item $[[iP → iN ≈ iQ → iM]]$\\
      Then by inversion of \ruleref{\ottdruleDOneArrowLabel},
      $[[iP ≈ iQ]]$ and $[[iN ≈ iM]]$. By the induction hypothesis,
      $[[Γ ⊢ iP ≈ iQ]]$ and $[[Γ ⊢ iN ≈ iM]]$, which means by inversion:
      \begin{enumerate*}
        \item[(i)] $[[Γ ⊢ iP ≥ iQ]]$,
        \item[(ii)] $[[Γ ⊢ iQ ≥ iP]]$,
        \item[(iii)] $[[Γ ⊢ iN ≤ iM]]$,
        \item[(iv)]  $[[Γ ⊢ iM ≤ iN]]$.
      \end{enumerate*}
      Applying \ruleref{\ottdruleDOneArrowLabel} to (i) and (iii), we obtain
      $[[Γ ⊢ iP → iN ≤ iQ → iM]]$; applying it to (ii) and (iv), we have $[[Γ ⊢
      iQ → iM ≤ iP → iN]]$. Together, it implies $[[Γ ⊢ iP → iN ≈ iQ → iM]]$.
    \item $[[∀pas.iN ≈ ∀pbs.iM]]$\\
      Then by inversion, there exists bijection $[[mu : ({pbs} ∩ fv iM) ↔ ({pas}
      ∩ fv iN)]]$, such that $[[iN ≈ [mu] iM]]$. By the induction hypothesis,
      $[[Γ, pas ⊢ iN ≈ [mu] iM]]$. From \cref{corollary:subst-pres-equiv} and
      the fact that $[[mu]]$ is bijective, we also have
      $[[Γ, pbs ⊢ [mu-1]iN ≈ iM]]$.

      Let us construct a substitution $[[{pas} ⊢ iPs/pbs : {pbs}]]$ by
      extending $[[mu]]$ with arbitrary positive types on $[[{pbs} \ fv iM]]$.

      Notice that $[[ [mu]iM ]] = [[ [iPs/pbs]iM ]]$, and therefore,
      $[[Γ, pas ⊢ iN ≈ [mu] iM]]$ implies $[[Γ, pas ⊢ [iPs/pbs]iM ≤ iN]]$. Then by
      \ruleref{\ottdruleDOneForallLabel}, $[[Γ ⊢ ∀pbs.iM ≤ ∀pas.iN]]$.

      Analogously, we construct the substitution from $[[mu-1]]$, and use it to
      instantiate $[[pas]]$ in the application of
      \ruleref{\ottdruleDOneForallLabel} to infer $[[Γ ⊢ ∀pas.iN ≤ ∀pbs.iM]]$.

      This way, $[[Γ ⊢ ∀pbs.iM ≤ ∀pas.iN]]$ and $[[Γ ⊢ ∀pas.iN ≤ ∀pbs.iM]]$
      gives us $[[Γ ⊢ ∀pbs.iM ≈ ∀pas.iN]]$.

    \item For the cases of the positive types, the proofs are symmetric.
  \end{caseof}
\end{proof}

\begin{corollary}[Normalization is sound w.r.t. subtyping-induced equivalence] 
  \label{corollary:nf-sound-wrt-subt-equiv}
  \hfill
  \begin{itemize}
    \item [$+$] if $[[Γ ⊢ iP]]$ then $[[Γ ⊢ iP ≈ nf(iP)]]$,
    \item [$-$] if $[[Γ ⊢ iN]]$ then $[[Γ ⊢ iN ≈ nf(iN)]]$.
  \end{itemize}
\end{corollary}
\begin{proof}
  Immediately from \cref{lemma:normalization-soundness,corollary:wf-nf,lemma:equiv-soundness}.
\end{proof}

\begin{corollary}[Normalization preserves subtyping] 
  \label{corollary:nf-pres-subt}
  Assuming all the types are well-formed in context $[[Γ]]$,
  \begin{itemize}
    \item [$+$] $[[Γ ⊢ iP ≥ iQ]] \iff [[Γ ⊢ nf(iP) ≥ nf(iQ)]]$,
    \item [$-$] $[[Γ ⊢ iN ≤ iM]] \iff [[Γ ⊢ nf(iN) ≤ nf(iM)]]$.
  \end{itemize}
\end{corollary}
\begin{proof}
  \hfill
  \begin{itemize}
    \item [$+$]  
    \begin{itemize}
      \item [$\Rightarrow$] Let us assume $[[Γ ⊢ iP ≥ iQ]]$.
        By \cref{corollary:nf-sound-wrt-subt-equiv},
        $[[Γ ⊢ iP ≈ nf(iP)]]$ and $[[Γ ⊢ iQ ≈ nf(iQ)]]$, 
        in particular, by inversion, 
        $[[Γ ⊢ nf(iP) ≥ iP]]$ and $[[Γ ⊢ iQ ≥ nf(iQ)]]$.
        Then by the transitivity of subtyping 
        (\cref{corollary:subtyping-transitivity}), 
        $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
      \item [$\Leftarrow$] Let us assume $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
        Also by \cref{corollary:nf-sound-wrt-subt-equiv}
        and inversion, 
        $[[Γ ⊢ iP ≥ nf(iP)]]$ and $[[Γ ⊢ nf(iQ) ≥ iQ]]$.
        Then by the transitivity, $[[Γ ⊢ iP ≥ iQ]]$.
    \end{itemize}
    \item [$-$] The negative case is proved symmetrically.
  \end{itemize}
\end{proof}

\begin{lemma}[Subtyping induced by disjoint substitutions]
  \label{lemma:subt-ind-disj-subst}
  If two disjoint substitutions induce subtyping, they are degenerate (so is the
  subtyping).
  Suppose that $[[Γ ⊢ σ1 : Γ1]]$ and $[[Γ ⊢ σ2 : Γ1]]$,
  where $[[Γi ⊆ Γ]]$ and $[[Γ1 ∩ Γ2= ∅]]$. Then
  \begin{itemize}
  \item[$-$] assuming $[[Γ ⊢ iN]]$,~
    $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ implies $[[Γ ⊢ σi ≈ id :  fv iN]]$
  \item[$+$] assuming $[[Γ ⊢ iP]]$,~
    $[[Γ ⊢ [σ1]iP ≥ [σ2]iP]]$ implies $[[Γ ⊢ σi ≈ id :  fv iP]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Proof by induciton on $[[Γ ⊢ iN]]$ (and mutually on $[[Γ ⊢ iP]]$).
  \begin{caseof}
    \item $[[iN]] = [[α⁻]]$\\
      Then $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ is rewritten as $[[Γ ⊢ [σ1]α⁻ ≤ [σ2]α⁻]]$.
      Let us consider the following cases:
      \begin{caseof}
      \item $[[α⁻ ∉ Γ1]]$ and $[[α⁻ ∉ Γ2]]$ \label{case:var-not-in-ctxts}\\
        Then $[[Γ ⊢ σi ≈ id : {α⁻}]]$ holds immediately,
        since $[[ [σi] α⁻]] = [[ [id] α⁻]] = [[α⁻]]$ and
        $[[Γ ⊢ α⁻ ≈ α⁻]]$.
      \item $[[α⁻ ∊ Γ1]]$ and $[[α⁻ ∊ Γ2]]$\\
        This case is not possible by assumption: $[[Γ1 ∩ Γ2= ∅]]$.
      \item $[[α⁻ ∊ Γ1]]$ and $[[α⁻ ∉ Γ2]]$\\
        Then we have $[[Γ ⊢ [σ1]α⁻ ≤ α⁻]]$,
        which by \cref{corollary:vars-no-proper-subtypes} means $[[Γ ⊢ [σ1]α⁻ ≈ α⁻]]$,
        and hence, $[[Γ ⊢ σ1 ≈ id : {α⁻}]]$.

        $[[Γ ⊢ σ2 ≈ id : {α⁻}]]$ holds since $[[ [σ2]α⁻ ]] = [[α⁻]]$,
        similarly to \cref{case:var-not-in-ctxts}.

      \item $[[α⁻ ∉ Γ1]]$ and $[[α⁻ ∊ Γ2]]$\\
        Then we have $[[Γ ⊢ α⁻ ≤ [σ2]α⁻]]$,
        which by \cref{corollary:vars-no-proper-subtypes} means $[[Γ ⊢ α⁻ ≈ [σ2]α⁻]]$,
        and hence, $[[Γ ⊢ σ2 ≈ id : {α⁻}]]$.

        $[[Γ ⊢ σ1 ≈ id : {α⁻}]]$ holds since $[[ [σ1]α⁻ ]] = [[α⁻]]$,
        similarly to \cref{case:var-not-in-ctxts}.
      \end{caseof}
  \item $[[iN]] = [[∀pas.iM]]$\\
    Then by inversion, $[[Γ, pas ⊢ iM]]$.
    $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ is rewritten as $[[Γ ⊢ [σ1]∀pas.iM ≤ [σ2]∀pas.iM]]$.
    By the congruence of substitution and by the inversion of
    \ruleref{\ottdruleDOneForallLabel}, $[[Γ, pas ⊢ [iQs/pas][σ1]iM ≤ [σ2]iM]]$,
    where $[[Γ, pas ⊢ iQi]]$.
    Let us denote the (Kleisli) composition of $[[σ1]]$ and $[[iQs/pas]]$ as
    $[[σ1']]$, noting that $[[Γ, pas ⊢ σ1' : Γ1, pas]]$,
    and $[[(Γ1, pas) ∩ Γ2 = ∅]]$.

    Let us apply the induction hypothesis to $[[iM]]$ and the
    substitutions $[[σ1']]$ and $[[σ2]]$ with
    $[[Γ, pas ⊢ [σ1']iM ≤ [σ2]iM]]$ to obtain:
    \begin{align}
      [[Γ, pas ⊢ σ1' ≈ id :  fv iM]] \label{fact:subs-proper-sub:forall-ih}\\
      [[Γ, pas ⊢ σ2 ≈ id :  fv iM]]  \label{fact:subs-proper-sub:forall-ih2}
    \end{align}

    Then $[[Γ ⊢ σ2 ≈ id :  fv ∀pas.iM]]$ holds by strengthening of
    \ref{fact:subs-proper-sub:forall-ih2}:
    for any $[[β±]] \in [[fv ∀pas.iM]] = [[fv iM \ {pas}]]$,
    $[[Γ, pas ⊢ [σ2]β± ≈ β±]]$ is strengthened to $[[Γ ⊢ [σ2]β± ≈ β±]]$, because
    $[[fv [σ2]β±]] = [[fv β±]] = \{[[β±]]\} \subseteq [[Γ]]$.

    To show that $[[Γ ⊢ σ1 ≈ id :  fv ∀pas.iM]]$, let us take an arbitrary
    $[[β±]] \in [[fv ∀pas.iM]] = [[fv iM \ {pas}]]$.

    $
    \begin{aligned}[t]
      [[β±]] &= [[ [id]β± ]]
             && \text{by definition of $[[id]]$}\\
             &\eqDOne [[ [σ1']β± ]]
             && \text{by \ref{fact:subs-proper-sub:forall-ih}}\\
             &= [[ [iQs/pas][σ1]β±]]
             && \text{by definition of $[[σ1']]$}\\
             &= [[ [σ1]β± ]]
             && \text{because $[[{pas} ∩ fv [σ1]β±]] \subseteq [[{pas} ∩ Γ]] = \emptyset$}
    \end{aligned}
    $\\
    This way, $[[Γ ⊢ [σ1]β± ≈ β±]]$ for any $[[β±]] \in [[fv ∀pas.iM]]$ and thus,
    $[[Γ ⊢ σ1 ≈ id :  fv ∀pas.iM]]$.

>>>>

@@

\begin{lemma}[Algorithmic type well-formedness is invariant under equivalence]
  \label{lemma:wf-equiv-algo}
  Mutual subtyping implies declarative equivalence.
  \begin{itemize}
  \item[$+$] if $[[uP ≈ uQ]]$ then $[[Γ; Ξ ⊢ uP]] \iff [[Γ; Ξ ⊢ uQ]]$,
  \item[$-$] if $[[uN ≈ uM]]$ then $[[Γ; Ξ ⊢ uN]] \iff [[Γ; Ξ ⊢ uM]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  The proof coincides with the proof of \cref{lemma:wf-equiv},
  and adds two cases for equating two positive or two negative algorithmic variables,
  which must be equal by inversion, and thus, 
  $[[Γ; Ξ ⊢ α̂±]] \iff [[Γ; Ξ  ⊢ α̂±]]$ holds trivially.
\end{proof}

@@

@@

\begin{lemma}[Algorithmic substitution equivalence becomes equality after normalization]
  \label{corollary:subst-subt-equiv-algorithmization}
  Suppose that $[[Θ ⊢ uσ1 : Ξ']]$ and $[[Θ ⊢ uσ2 : Ξ']]$ are algorithmic substitutions
  and $[[Ξ ⊆ Ξ']]$. 
  Then $[[Θ ⊢ uσ1 ≈ uσ2 : Ξ]] \iff [[nf(uσ1)|Ξ = nf(uσ2)|Ξ]]$.
\end{lemma}
\begin{proof}
  Follows immediately from \cref{lemma:subt-equiv-algorithmization}:
  \begin{itemize}
    \item [$\Rightarrow$]
      If $[[α̂± ∉ Ξ]]$, then $[[ [nf(uσ1)|Ξ]α̂± = [nf(uσ2)|Ξ]α̂± ]] = [[α̂±]]$ by 
      definition. 
      For any $[[α̂± ∊ Ξ]]$, 
      $[[ [nf(uσ1)|Ξ]α̂± = nf([uσ1]α̂±)]]$ and 
      $[[ [nf(uσ2)|Ξ]α̂± = nf([uσ2]α̂±) ]]$; 
      $[[Θ(α̂±) ⊢ [uσ1]α̂± ≈ [uσ2]α̂±]]$ implies
      $[[ nf([uσ1]α̂±) = nf([uσ2]α̂±) ]]$ by \cref{lemma:decl-equiv-algorithmization}.
    \item [$\Leftarrow$]
      If $[[α̂± ∊ Ξ]]$, then 
      $[[nf(uσ1)|Ξ = nf(uσ2)|Ξ]]$ implies
      $[[ nf([uσ1]α̂±) = nf([uσ2]α̂±) ]]$ by definition 
      of substitution restriction and normalization.
      In turn, $[[ nf([uσ1]α̂±) = nf([uσ2]α̂±) ]]$ means
       $[[Θ(α̂±) ⊢ [uσ1]α̂± ≈ [uσ2]α̂±]]$ by \cref{lemma:decl-equiv-algorithmization}.
  \end{itemize}
\end{proof}
